---
title: "CH0 Basic R"
output: pdf_document
date: "2022-09-04"
---

I. Data Types
  info stored in R can be in diff types 
  4 types:
    1. numeric (contains double/floating): 2.33  1  pi
    2. integer: 10L 100L (enter L to let R know we are having an integer)
    3. logical: TRUE FALSE T F
    4. character: 'a string text'  '2.33'
```{r}
a = 10L; b = 10; c = '10'; d = T
class(a); class(b); class(c); class(d)
```

II. Data Structures
  ways data be organized
  4 types:
    1. vector - 1 dimension 1 type of data
    2. list - 1 dimension mix types of data
    3. matrix - 2 dimension 1 type of data
    4. data frame - 2 dimension mix types of data
    
## 1.1 Vectors
```{r}
# numeric vector
a1 = c(1,2,3,4:7)
class(a1); a1
a2 = c(1,2.33,3,7:4) # every elements auto floating
class(a2); a2
a3 = seq(1,10,2)
class(a3); a3

# extensions for floating vs integers
a1.flo = as.double(a1)
a1; a1.flo; class(a1.flo)
a1.realflo = format(round(a1,2),nsmall=2)
a1.realflo;class(a1.realflo)
cat(a1.realflo)

a2.asint= as.integer(a2) # 2.33 automatically be rounded into int
a2; a2.asint

# character vector
# cannot use b = 3*'you are beautiful'
b = c('x','y','z')
class(b); print(b)

# extension: how to repeat characters? use functions
c1 = rep('You are Beautiful',3) # avoid using c as name since c() is a build-in function name
class (c1); c1; length(c1)
d = strrep('You are Beautiful ',3) # strrep() repeat and concatenate
class(d); d; length(d)

# If mix
mix1 = c(1,2,'hi')
mix1; class(mix1)
mix2 = c(T,F,1)
mix2; class(mix2)
mix3 = c(T,F,1,'hi')
mix3; class(mix3)
```
## 1.2 Factors
Factors are variables in R that often refereed as categorical variables.
Factors are stored as a vector of integer values with characters.
```{r}
data1 = c(1,2,2,3,1,2,3,3,1,2,3,3,1)
f.data1 = factor(data1)
f.data1 # R auto gives levels # f.data1 left out print(), they are the same thing
cat('the class is: ',class(f.data1))

# change 1,2,3 to I, II, III can directly change level, can be used in encrypting
levels(f.data) = c('I','II','III')
f.data

# example, election:
c.election = c('Biden','Trump','Biden','Trump','Hawkins','Biden','Jorgensen','Biden')
table(c.election)
c.election.factor = factor(c.election)
c.election.factor
# make the listing order in Trump < Biden < Hawkins < Jorgensen
c.election = factor(c.election,levels = c('Trump','Biden','Hawkins','Jorgensen'),ordered = T)
# table and summary give the same thing, str() notify it is a factor and gives some basic info
table(c.election)
summary(c.election)
str(c.election)
# levels list all the levels
levels(c.election)
factor(c.election)
```
## 2. List; one dimensional and hold multiple data types
   one list can include another list
```{r}
list1 = list('A',1,T)
class(list1)
names(list1)
list1
list1[1]


mix.list = list(chara = c('You','are'),
                vec = c(1,2.33),
                a.matrix = cbind(c(1, 2), c(3, 4)),
                a.list = list(c(5,6.3), c('Beautiful', '!')))
names(mix.list)
mix.list
mix.list[4]

```
## 3. Matrix; 2 dimensional same data type
  Two ways to create matrix: 
    1. use matrix() directly or 
    2. combine vectors, 1 vec for 1 col using cbind(vec1,vec2,...); 1 vec for 1 row using rbind(row1,row2,...) 
```{r}
# Way One using matrix()
matrix1 = matrix(data = c(1,2,3,4), 
                 nrow = 2, # assign num of rows
                 ncol = 2, # assign num of columns
                 byrow = TRUE) # default is FALSE which is assigning the num vertically, 
                             # by assigning byrow as TRUE, read numbers horizontally

matrix1.simple = matrix (c(1,2,3,4),2,2)
matrix1; matrix1.simple

dim(matrix1) # show dimensions: nrow, ncol

# Way Two using cbind(), rbind()
vec1 = c(1,2,3)
vec2 = c('Cat','Dog','Bird')
vec3 = c('Cute','Dumb','no offense' )

animal.matrix = cbind(vec1,vec2,vec3)
animal.matrix # all the elements are transferred into characters now since matrix can only store one type of data
```

## 4. Data Frames; two dimensions and mix types of data
   using data.frame(col1,col2,col3,...) ; colN signed as colName = c(Xij) 
```{r}
df1 = data.frame(Rank = c(1,2,3),
                 Kinds = c('Cat','Dog','Bird'),
                 Comments = c('Cute','Dumb','no offense'))

df1
dim(df1)
names(df1)
ncol(df1)

# simple practice as a data science
# Download ISLR if you do not have it.
# you need to go to Tools, Install Packages, and type ISLR in packages
library(ISLR)
#head(Auto)
#summary(Auto)

# draw a plot
plot(Auto$mpg,Auto$displacement) # dot plot
lines(lowess(Auto$mpg,Auto$displacement)) # draw the trends line using lines(lowess(x,y))
```

III. Operators
  how things works
  4 types:
    1. assignment operator
    2. arithmetic operators
    3. relational operators
    4. logical operators
```{r}

```

IV. Sub-setting Data Structures
  extract elements from a data structure
  3 types:
    1. sub-setting vectors
    2. sub-setting lists
    3. sub-setting data frames
```{r}

```

    
